---
title: "No more dividing likelihoods !"
author: "Loecher"
date: "2026-05-20"
output: 
  html_document:
    toc: True
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = F, warning = F, cache = F)

library(ranger)
library(mlr3learners)
library(arf)
library(ggpubr)
library(ggplot2)

source("utils.R")
genData = F
set.seed(123)
n_samples = 10000
num_trees = 200L
min_node_size = 20
```

## Systematic Comparison via DGP

The goal is benchmarking ARF against a ground-truth Titanic simulation, 
Here's a DGP with a distinct distribution for each group:

### What each group stress-tests in ARF

| Group | Distribution | ARF challenge |
|---|---|---|
| female/1 | Bimodal Gaussian | Valley between modes on linear scale |
| male/1 | Bimodal Log Normal | Very difficult
| female/2 | Power law (1/x²) | Extreme skew, mass near lower bound |
| male/2 | Uniform | Flat density — ARF leaf Gaussians will over-smooth edges |
| female/3 | Gamma | Moderate skew, unimodal — ARF's 2nd easiest case |
| male/3 | Spike + gamma tail |  very different scales |
  
  
The uniform and power law cases are particularly interesting diagnostics — ARF's truncated normal leaves will visibly over-smooth the sharp edges of the uniform, and will struggle with the extreme mass concentration near £5 in the power law.


Global Parameters:

- N = `r n_samples`
- n_trees = `r num_trees`
- min_node_size = `r min_node_size`


```{r}
if (genData) 
  sim_data <- as.data.table(simulate_titanic_dgp(n = n_samples, seed = NULL))[, -"male"]
#hist(sim_data[Pclass == 1 & sex == "female", Fare], nclass = 30)
#hist(syn_data$Fare, nclass = 30)
```

```{r, eval=F,echo=F}
library(ggplot2)
x1 = cbind.data.frame(Fare = sim_data[Pclass == 1 & sex == "female", Fare], type= "original", Pclass = 1, sex = "female")
x2 = cbind.data.frame(Fare = syn_data$Fare, type= "synthetic",  Pclass = 1, sex = "female")
x = rbind.data.frame(x1,x2)
ggplot(x, aes(x=Fare, fill = type)) + geom_histogram(alpha=0.5) #+ facet_wrap(~ type)
```

```{r}

CondDensityEstimator = function(mtry=2, lik_grid = 300, 
                                num_trees = 100,
                                min_node_size = 20,
                                finite_bounds = "local",
                                arf = NULL){
  if (is.null(arf)) arf <- adversarial_rf(sim_data, mtry = mtry, num_trees = num_trees, min_node_size = min_node_size)
  
  params <- forde(arf, sim_data, finite_bounds = finite_bounds)

  allData = allLiks = NULL
  for (p in 1:3){
    for (s in c("female", "male")){
      syn_data <- forge(params, evidence = data.table(Pclass = p, sex = s),
                 n_synth = sim_data[Pclass == "3", .N])
      x1 = cbind.data.frame(Fare = sim_data[Pclass == p & sex == s, Fare], type= "original", Pclass = p, sex = s)
      x2 = cbind.data.frame(Fare = syn_data$Fare, type= "synthetic",  Pclass = p, sex = s)
      x = rbind.data.frame(x1,x2)
      allData = rbind.data.frame(allData,x)
      ################################################
      if (lik_grid > 0){ # also compute the likelihoods and densities
        psi = params
        fares  <- exp(seq(log(1), log(300), length.out = lik_grid))
        #browser()
        ll <- lik(psi, data.table(Fare = fares) , arf = arf, evidence = data.table(Pclass = p, sex = s), log = TRUE)
        l_syn <- cbind.data.frame(Fare = fares, likelihood = exp(ll), type= "synthetic",  Pclass = p, sex = s)
        l_orig = cbind.data.frame(Fare = fares, likelihood = get_cond_density(fares,pclass = p, sex = s), type= "original",  Pclass = p, sex = s)
        
        l = rbind.data.frame(l_syn, l_orig)
        allLiks = rbind.data.frame(allLiks,l)
      }
    }
  }

  return(list(hist = allData, lik = allLiks, arf = arf))
}


 plotHists = function(allData, commonAxis = TRUE){
   if (commonAxis){
     g = ggplot(allData, aes(x=Fare, y=after_stat(density), fill = type)) + geom_histogram(alpha=0.5,
                 position="identity") + facet_wrap(~ Pclass + sex, scales = "free")
   } else {
     #heavy tails: female/2 male/3 male/1
     idx = (allData$Pclass == 2 & allData$sex == "female") | (allData$Pclass == 3 & allData$sex == "male") | (allData$Pclass == 1 & allData$sex == "male")
     LightTails = allData[!idx,]
     HeavyTails = allData[idx,]
     g1 = ggplot(LightTails, aes(x=Fare, y=after_stat(density), fill = type)) + geom_histogram(alpha=0.5,
                 position="identity") + facet_wrap(~ Pclass + sex, scales = "free") 
     g2 = ggplot(HeavyTails, aes(x=Fare, y=after_stat(density), fill = type)) + geom_histogram(alpha=0.5,
                 position="identity") + facet_wrap(~ Pclass + sex, scales = "free") + scale_x_log10()
     g = ggarrange(g1,g2, ncol=1, labesl = c("Linear scale", "Log scale"))
   }
  
   return(g) 
 }

 plotLiks = function(allData, commonAxis = TRUE){
   allData$type = factor(allData$type, levels = c("synthetic", "original"))
   if (commonAxis){
     g = ggplot(allData, aes(x=Fare, y=likelihood, colour  = type)) + geom_line(linewidth = 0.9) + facet_wrap(~ Pclass + sex, scales = "free")
   } else {
     #heavy tails: female/2 male/3
     idx = (allData$Pclass == 2 & allData$sex == "female") | (allData$Pclass %in% c(1,3) & allData$sex == "male")
     LightTails = allData[!idx,]
     HeavyTails = allData[idx,]
     g1 = ggplot(LightTails, aes(x=Fare, y=likelihood, colour = type)) + geom_line(linewidth = 0.9) + facet_wrap(~ Pclass + sex, scales = "free") 
     g2 = ggplot(HeavyTails, aes(x=Fare, y=likelihood, colour = type)) + geom_line(linewidth = 0.9)+ facet_wrap(~ Pclass + sex, scales = "free") + scale_x_log10()
     g = ggarrange(g1,g2, ncol=1, labesl = c("Linear scale", "Log scale"))
   }
  
   return(g) 
 }
```

# finite_bounds = "local"

This seems to make a huge difference

```{r genData}
if (genData){
  print("estimating ARFs")
  allData1 = CondDensityEstimator(1,num_trees = num_trees, min_node_size=min_node_size);print("mtry 1 done")
  allData2 = CondDensityEstimator(2,num_trees = num_trees, min_node_size=min_node_size);print("mtry 2 done")
  allData3 = CondDensityEstimator(3,num_trees = num_trees, min_node_size=min_node_size);print("mtry 3 done")
  save.image("allData.rda")
} else {
  genDataOrig = genData
  load("allData.rda")
  genData = genDataOrig
}
```

## mtry = 1

This is effectively randomized trees.

### Data Generation

```{r hist1, warning = FALSE, message = FALSE, results = F, fig.width=10, fig.height=8}
g1 = plotHists(allData1$hist,F)
print(g1)
```

### Likelihood Densities


```{r lik1, warning = FALSE, message = FALSE, fig.width=10, fig.height=8}
g1 = plotLiks(allData1$lik,F)
print(g1)
```
## Forest Diagnostics

.e.g how big are the leafs ?


## mtry = 2

### Data Generation

```{r hist2, warning = FALSE, message = FALSE, fig.width=10, fig.height=8}
g2 = plotHists(allData2$hist,F)
print(g2)
```

### Likelihood Densities


```{r lik2, warning = FALSE, message = FALSE, fig.width=10, fig.height=8}
g2 = plotLiks(allData2$lik,F)
print(g2)
```

## mtry = 3

This is effectively bagging.

### Data Generation

```{r hist3, warning = FALSE, message = FALSE, fig.width=10, fig.height=8}
g3 = plotHists(allData3$hist,F)
print(g3)
```

### Likelihood Densities


```{r lik3, warning = FALSE, message = FALSE, fig.width=10, fig.height=8}
g3 = plotLiks(allData3$lik,F)
print(g3)
```


----------------------------------------

# default finite_bounds ("no")

This was my original try

```{r genData_no}
if (genData){
  print("estimating ARFs")
  allData1_no = CondDensityEstimator(1,num_trees = num_trees, min_node_size=min_node_size, finite_bounds = "no");print("mtry 1 done")
  allData2_no = CondDensityEstimator(2,num_trees = num_trees, min_node_size=min_node_size, finite_bounds = "no");print("mtry 2 done")
  allData3_no = CondDensityEstimator(3,num_trees = num_trees, min_node_size=min_node_size, finite_bounds = "no");print("mtry 3 done")
  save.image("allData.rda")
} else {
  genDataOrig = genData
  load("allData.rda")
  genData = genDataOrig
}
```

## mtry = 1

This is effectively randomized trees.

### Data Generation

```{r hist1_no, warning = FALSE, message = FALSE, results = F, fig.width=10, fig.height=8}
g1 = plotHists(allData1_no$hist,F)
print(g1)
```

### Likelihood Densities


```{r lik1_no, warning = FALSE, message = FALSE, fig.width=10, fig.height=8}
g1 = plotLiks(allData1_no$lik,F)
print(g1)
```

## mtry = 2

### Data Generation

```{r hist2_no, warning = FALSE, message = FALSE, fig.width=10, fig.height=8}
g2 = plotHists(allData2_no$hist,F)
print(g2)
```

### Likelihood Densities


```{r lik2_no, warning = FALSE, message = FALSE, fig.width=10, fig.height=8}
g2 = plotLiks(allData2_no$lik,F)
print(g2)
```

## mtry = 3

This is effectively bagging.

### Data Generation

```{r hist3_no, warning = FALSE, message = FALSE, fig.width=10, fig.height=8}
g3 = plotHists(allData3_no$hist,F)
print(g3)
```

### Likelihood Densities


```{r lik3_no, warning = FALSE, message = FALSE, fig.width=10, fig.height=8}
g3 = plotLiks(allData3_no$lik,F)
print(g3)
```


----------------------------------------------

# Conclusions

* Uniform density seems hardest
* Higher `mtry` -> better fit initially followed by "overfitting" of the likelihood.
* The extreme lack of smoothness of the likelihoods is not really visible in the histograms.
* Using the `evidence` argument in `lik()` is MUCH superior to dividing by the marginal likelihoods
  * With hindsight it seems clear why: the marginal likelihoods are likely (no pun) also very noisy; dividing by small, noisy numbers will lead to amplified noise.
* `finite_bounds` does not seem to matter much after all

# Outlook

* Can we efficiently and soundly smooth the likelihoods ? Could hierarchical shrinkage play a role ? (Just increasing `num_trees` does not seem to be enough)
* How can one better approximate a uniform distribution ? Maybe expand the local basis functions ?
* Can we tune mtry ?
* 



```{r}
#if (genData) save.image("allData.rda")
```

